STAT540 Seminar 01: Exploration of small gene expression dataset
---
_Gene expression data from photo receptor cells in mice_  
_Rebecca Johnston 11/01/2014_

Here I will be exploring _prDat_, a gene expression dataset from photo recptor
cells in mice, from various developmental stages and two genotypes. Note we will 
only work with gene expression data for three genes, and their names have been 
replaced with 3 Pokemon attacks (crabHammer, eggBomb, and poisonFang).

* <dim id="1a">[Load GSE4051\_MINI.txt _prDat_](#1b)
* <dim id="2a">[Basic exploration of _prDat_](#2b)  
* <dim id="3a">[Indexing and subsetting _prDat_](#3b)


### <dim id="1b">[Load GSE4051\_MINI.txt _prDat_](#1b)

Save GSE4051_MINI.txt to wd from [STAT540 github](https://github.com/jennybc/stat540_2014/tree/master/examples/photoRec/data)

```{r}
prDat <- read.table("GSE4051_MINI.txt", header = TRUE, row.names = 1)
```


### <dim id="2b">[Basic exploration of _prDat_](#2b)  

```{r}
str(prDat) # Data structure
summary(prDat) # Data summary
nrow(prDat) # How many rows are there?
ncol(prDat) # How many columns or variables are there?
head(prDat) # Inspect the first few observations
tail(prDat) # Or the last few
prDat[5:16, ] # Or a random sample
prDat[sample(nrow(prDat), size = 6), ] # Random sample using Jenny's code! Cool!
```


QUESTION: What does row correspond to -- different genes or different mice?  
ANSWER: Rows correspond to different mice (samples).  

```{r}
names(prDat) # What are the variable names?
str(prDat) # What "flavor" is each variable, i.e. numeric, character, factor?
```

QUESTION: For "sample", does each integer between 1 and nrows occur exactly once?
```{r}
nrow(prDat) # length(prDat$sample)
table(prDat$sample)
sort(prDat$sample) # Sort sample numbers in ascending order
seq(1, nrow(prDat)) # Create sequence of numbers from 1 to nrow in dataset
sort(prDat$sample) == seq(1, nrow(prDat)) # Determine if 2 variables are equal
all.equal(sort(prDat$sample), seq(1, nrow(prDat)))
identical(sort(prDat$sample), seq(1, nrow(prDat)))
```

ANSWER: Yes, each integer between 1 and nrows occur exactly once for "sample".  

```{r}
levels(prDat$devStage) # For each factor variable, what are the levels? 
levels(prDat$gType)
summary(prDat$devStage) # How many observations for each level of devStage?
summary(prDat$gType) # For gType?
table(prDat$devStage, prDat$gType) # Perform a cross-tabulation of devStage and gType
```

QUESTION: What do you think the intended experimental design was? What actually happened in real life?  
ANSWER: From the cross-tabulation above of genotype (gType) and developmental stage (devStage), I would assume that the intended experimental design was to compare two different genotypes in mice (wild type and knock out) across 5 different developmental stages. There were meant to be 4 mice for each of the 5 developmental stages, and 4 mice for each of the two genotypes, but it appears in "real life", one of the knock out mice (NrlKO) at developmental stage E16 was not able to be obtained.  


QUESTION: For each quantitative variable, what are the extremes? How about average or median?  
ANSWER: Use summary for the variable names we want

```{r}
summary(prDat[ , c("crabHammer", "eggBomb", "poisonFang")])
range(prDat$crabHammer) # returns min and max
fivenum (prDat$crabHammer) # returns Tukey's five number summary
```


### <dim id="3b">[Indexing and subsetting _prDat_](#3b)

Create a new data.frame called weeDat only containing observations for which expression of poisonFang is above 7.5
```{r}
prDat$poisonFang > 7.5
weeDat <- prDat$poisonFang[which(prDat$poisonFang > 7.5)] # Use "which" argument and indexing
str(weeDat)
```

For how many observations poisonFang > 7.5? How do they break down by genotype and developmental stage?
```{r}
length(weeDat)
prDat[prDat$poisonFang %in% weeDat, ]
prDat[prDat$poisonFang %in% weeDat, c("poisonFang", "gType", "devStage")]
prDat[prDat$poisonFang > 7.5, ]
```

Print the observations with row names "Sample\_16" and "Sample\_38" to screen, showing only the 3 gene expression variables.
```{r}
prDat[c("Sample_16", "Sample_38"), c("crabHammer", "eggBomb", "poisonFang")]
```

Which samples have expression of eggBomb less than the 0.10 quantile?
```{r}
sort(prDat$eggBomb)
quantile(prDat$eggBomb, probs = 0.1) # Answer = 6.1844
# quantile does not return a number. Can you code this instead of quoting it?
prDat[prDat$eggBomb < 6.1844, c("eggBomb", "sample")]
# Yes you can! Just add the two lines together and voila! Jenny's code:
rownames(prDat[prDat$eggBomb < quantile(prDat$eggBomb, 0.1), ])
```
